\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-seq data\\using R packages}}

### 2012-01-10 package/smutans/inst/doc/smutans.Rnw
R package smutans_0.1.3

\SweaveOpts{keep.source=TRUE,eps=FALSE,include=FALSE,width=4,height=4.5}
\newcommand{\Robject}[1]{object \texttt{#1}}
\newcommand{\Rpackage}[1]{R package \textit{#1}}
\newcommand{\Rclass}[1]{class \textit{#1}}
\newcommand{\Rfunction}[1]{{function \small\texttt{#1}}}
\newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}

@@

\title{\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-seq data using R packages}}}

@@

\begin{abstract}
We analyze a number of RNA samples from \textit{Streptococcus mutans}.  We use
RNA-seq to define genes whose expression changes between samples
\cite{Oshlack2010}. Short reads are first mapped to a reference genome using
approaches such as BWA \cite{Li2009a}.  Reads fitting into one of the annotated
intervals are counted.  These counts are passed to a statistical package which
treats differences in read counts for statistical significance. For this
purpose, we made use of \Rpackage{DEseq} \cite{Anders2010} or \Rpackage{edgeR}
\cite{Robinson2010}. This prodecure returns a list of differentially expressed
genes, which can be passed to downstream analysis, such as functional category
enrichment test using \Rpackage{goseq} \cite{Young2010}.  This
\Rpackage{smutans} contains procedures that make use of functions provided by
the above R packages. 
\end{abstract}

@@

Steps of analyzing RNA-seq data for determining differentially expressed genes
are well described in \cite{Oshlack2010}. We used a short read aligner called
BWA \cite{Li2009a} to map the sequenced short reads on a reference genome,
\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
aligner allowed a few gaps for efficient alignment of millions of sequences of
size being 100 base pairs long, short reads that consist of RNAs and sequencing
adapter sequences would be not mapped. We used cutadapt \cite{Martin2011} to
remove parts of adapter sequences from short reads before mapping on the
reference genome. Mapped short read alignments were converted into readable
formats \cite{Li2009b}. We counted short reads aligned to annotated genes in the
reference genome, forming a table of read counts of the genes. Statistical
software R \cite{R2011} with its one of R package called DEseq \cite{Anders2010}
was subsequently employed to infer differentially expressed in different
biological conditions.  We also generated new tracks for our recently released
Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
summarize gene expression levels based the short read alignments. This allowed
us to visualize parts of the reference genome for comparing different RNA
samples. Further analyses such as functionaly category enrichment test
\cite{Young2010} were employed.

The \Rpackage{smutans} expects count data in the form of \Rclass{CountDataSet},
which is a type provided by \Rpackage{DESeq}.  Using BWA \cite{Li2009a} and
samtools \cite{Li2009b} we aligned short reads from RNA sequencing. Using gene
annotations in the UA159 genome we counted short reads that overlapped the
genes.  The count values are raw counts of sequencing reads. The
\Rpackage{DESeq}'s statistical model, or negative binomial model, hold for raw
reads that allows to assess the measurement precision correctly; we did not have
rounded values of normalized counts, or counts of covered base pairs.  Several
exeriments of \textit{S.\ mutans} RNA sampling were performed.  \fixme{Find the
papers that are related with the experiments.} One of them was presented in the
\Robject{smutansGenes} as a part of \Rpackage{smutans}. The
\Robject{smutansGenes} is of \Rclass{CountDataSet}, which is the data type used
by \Rpackage{DESeq}. We have a convenient \Rclass{Smutans}, which contains slots
for \Rclass{CountDataSet} objects, and results from analyses thereof.  Before
loading necessary R packages and data sets, let us describe what functions were
used for preparing data objects such as \Robject{smutansGenes}. One should
create an R script file at the directory that contains the source code directory
of the \Rpackage{smutans}. Put the following scripts in the R script file and
run the script at that directory. That should create RData files at the
directory smutans/data. User should not run the R script because the data files
should be found in the \Rpackage{smutans}.

>>>>

\SweaveOpts{keep.source=TRUE,eps=FALSE,include=FALSE,width=4,height=4.5}
\newcommand{\Robject}[1]{object \texttt{#1}}
\newcommand{\Rpackage}[1]{R package \textit{#1}}
\newcommand{\Rclass}[1]{class \textit{#1}}
\newcommand{\Rfunction}[1]{{function \small\texttt{#1}}}
\newcommand{\program}[1]{{\texttt{#1}}}
\newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}

@@

\title{\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-seq data\\using R packages}}}

@@

\begin{abstract}
We analyze RNA samples from \textit{Streptococcus mutans}.  We use
RNA-seq to define genes whose expression changes between samples
\cite{Oshlack2010}. Short reads are first mapped to a reference genome using
approaches such as \program{BWA} \cite{Li2009a}.  Reads fitting into one of the annotated
intervals are counted.  These counts are passed to a statistical package which
treats differences in read counts for statistical significance. For this
purpose, we made use of \Rpackage{DEseq} \cite{Anders2010} or \Rpackage{edgeR}
\cite{Robinson2010}. This prodecure returns a list of differentially expressed
genes, which can be passed to downstream analysis, such as functional category
enrichment test using \Rpackage{goseq} \cite{Young2010}.  The current
\Rpackage{smutans} contains procedures that use functions provided by
the above R packages. 
\end{abstract}

@@

Steps of analyzing RNA-seq data for determining differentially expressed genes
are described in \cite{Oshlack2010}. We used a short read aligner called
\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference genome,
\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
aligner allowed a few gaps for efficient alignment of millions of sequences of
size being 100 base pairs long, short reads that consist of RNAs and sequencing
adapter sequences would be not mapped. We used \program{cutadapt} \cite{Martin2011} to
remove parts of adapter sequences from short reads before mapping on the
reference genome. Mapped short read alignments were converted into readable
formats \cite{Li2009b}. We counted short reads aligned to annotated genes in the
reference genome, forming a table of read counts of the genes. Statistical
software R \cite{R2011} with \Rpackage{DEseq} \cite{Anders2010}
was subsequently employed to infer differentially expressed in different
biological conditions.  We also generated tracks for our recently released
Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
summarize gene expression levels based the short read alignments. This allowed
us to visualize parts of the reference genome for comparing different RNA
samples. Further analyses such as functional category enrichment test
\cite{Young2010} were also employed.

The \Rpackage{smutans} expects count data in the form of \Rclass{CountDataSet},
which is a type provided by \Rpackage{DESeq}.  Using \program{BWA} \cite{Li2009a} and
\program{samtools} \cite{Li2009b} we aligned short reads from RNA sequencing. Using gene
annotations in the UA159 genome we counted short reads that overlapped the
genes.  The count values are raw counts of sequencing reads. The
\Rpackage{DESeq}'s statistical model, or negative binomial model, hold for raw
reads that allows to assess the measurement precision correctly; we did not have
rounded values of normalized counts, or counts of covered base pairs.  Several
exeriments of \textit{S.\ mutans} RNA sampling were performed.  \fixme{Find the
papers that are related with the experiments.} One of them was presented in the
\Robject{smutansGenes} as a part of \Rpackage{smutans}. The
\Robject{smutansGenes} is of \Rclass{CountDataSet}, which is the data type used
by \Rpackage{DESeq}. We have a convenient \Rclass{Smutans}, which contains slots
for \Rclass{CountDataSet} objects, and results from analyses thereof.  Before
loading necessary R packages and data sets, let us describe what functions were
used for preparing data objects such as \Robject{smutansGenes}. One should
create an R script file at the directory that contains the source code directory
of the \Rpackage{smutans}. Put the following scripts in the R script file and
run the script at that directory. That should create RData files at the
directory smutans/data. User should not run the R script because the data files
should be found in the \Rpackage{smutans}.
